\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-31 doc/publications/plos16/plos16.tex
updated result section 2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
We investigate the ability of using a random forest model we build on genomic data to identify variants related to a
specific phenotype.  In this case we aim to identify variants associated with Bone Mineral Density (BMD). We build a
model on the BMD dataset (see datasets) using a binary high or low BMD value as the label.  From this model, we output a
list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with the highest Gini impurity,
and hence importance, is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.  Then, using the 26
known and 6 novel BMD genes from \cite{Duncan.2011}, we compile a list of variants that are present in or near one of
these 32 genes.

We then apply the Mann–Whitney U test to this list to verify that from our list, variants from known BMD loci are ranked
higher than variants from elsewhere in the genome.  We observe $P=1.3\times10^{-07}$, supporting our hypothesis that
variants in close proximity to the known BMD genes are contributing more information to our classification model.

However, we also observe that from the set of highest-ranked features, for example, the top ten, not all of them exist
within close proximity to known BMD genes. We investigate the literature to determine the plausibility of these variants
contributing to BMD.

One gene is COLEC10 (collectin 10). Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as
containing variants associated with BMD \cite{Kemp2014,Liu2008}. However there currently exists no information on how this gene
may contribute to BMD.

We also frequently see highly-ranked variants in chromosome 22q11. \cite{Stagi.2010}, observe that individuals with a
microdeletion in this region have a significant in bone mass\cite{Stagi2010}. In this same region, we also observe
variants present in PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as
proposed by \cite{Phang.2008}, proline is made available by collagen degradation \cite{Phang2008}.

DCDC5 (Doublecortin Domain Containing 5) is one of the 26 known genes, however it is one of the genes that \cite{Duncan.2011}.
failed to replicate in their study. DCDC5 does appear in our list of genes, however the first occurrence of a variant in
or near this gene has a rank of 9,667 out of 10,000. Also, this gene is not highly expressed in bone, and any mechanisms
by which this gene might regulate BMD are unclear \cite{Thakker2012}. Although variants near DCDC5 are associated with
BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
In this section we use \cursedforest\ to identify genomic locations that are associated with a specific trait. To do this, we perform feature selection over XX genomic variants to 
identify the locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. We build a
model on the BMD dataset (see datasets) using a binary high or low BMD value as the label.  From this model, we output a
list of the top 10,000 contributing variants, ranked by their Gini impurity. The variant with the highest Gini impurity,
and hence importance, is ranked 1, and the variant with the lowest Gini impurity is ranked 10,000.

%DENIS: number
To test whether the associated features have biological meaning, we investigate whether the known BMD genes investigated by Duncan {\it et al.}~\cite{Duncan.2011} are enriched amongst the 
highly ranked genes. We therefore label all variants that fall within XXbp of these 32 BMD genes and observe that they are
ranked significantly higher than other locations in the genome (Mann-Whitney U, $p=1.3\times10^{-07}$), supporting our hypothesis that biologically relevant features are contributing more information to our classification model.

We therefore investigate the highest-ranked features not amongst the associated gene list by Duncan {\it et al.}~\cite{Duncan.2011}. 
%DENIS is this correct about the transcription factor binding site?
COLEC10 (collectin 10) contains variants more recently associated with BDM~\cite{Kemp2014,Liu2008}, with a transcription factor binding site for the known BMD gene FOXL1 overlapping the highly ranked variant.
%However, we also observe that from the set of highest-ranked features, for example, the top ten, not all of them exist
%within close proximity to known BMD genes. We investigate the literature to determine the plausibility of these variants
%contributing to BMD.
%One gene is COLEC10 (collectin 10). Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as
%containing variants associated with BMD \cite{Kemp2014,Liu2008}. However there currently exists no information on how this gene
%may contribute to BMD. 
Also highly-ranked are variants in chromosome 22q11. Stagi {\it et al.}~\cite{Stagi.2010}, observe that individuals with a
microdeletion in this region have a significant reduction in bone mass. In this same region, we also observe
variants present in PRODH (proline dehydrogenase). PRODH catalyzes the first step in proline degradation, and as
proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is made available by collagen degradation.

Investigating the lowly-ranked genes, we observe DCDC5 (Doublecortin Domain Containing 5) ranked 9,667 out of 10,000. This gene is 
one of the BMD-associated genes postulated by literature that Duncan {\it et al.}~\cite{Duncan.2011} inspect.
Similar to their findings, our results also indicate that this gene does not have a strong association with BMD. 
In fact, DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear \cite{Thakker2012}. 
Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as to whether the gene itself contributes to BMD.
